---
title: "Assignment 1"
author: "Zakari L. Billo"
date: "Febrauary 23, 2026"
output: pdf_document
---

```{r setup, include=FALSE}
library(designmatch)
library(SuperLearner)
library(tmle)
library(kableExtra)
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Introduction

# Research Question

# Data 

```{r}
data(lalonde)


```

# Causal Estimand and Identification 

# Estimating the Causal Effect

We apply the three estimators discussed in class to determine the $\hat{ATE}$: g-computation (standardization), augmented inverse-propensity of treatment weights (AIPTW), and targeted maximum likelihood estimation with Ensemble SuperLearner (TMLE). 

# Comparing Approaches

# Conclusion

## References:

Dehejia, R., and Wahba, S. (1999), "Causal Effects in Nonexperimental Studies: Reevaluating the Evaluation of Training Programs," Journal of the American Statistical Association, 94, 1053-1062.

Laan, M. J. van der, Polley, E. C., & Hubbard, A. E. (2007). Super Learner. Statistical Applications in Genetics and Molecular Biology, 6(1). https://doi.org/10.2202/1544-6115.1309

Lalonde, R. (1986), "Evaluating the Econometric Evaluations of Training Programs," American Economic Review, 76, 604-620.

